---
title: "data wrangling into"
author: "Daniel Kerstan"
date: "8/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(palmerpenguins)
```

warm up
```{r}
us_rent_income_wider <- us_rent_income %>%  
  pivot_wider(names_from = variable, values_from = estimate) %>% 
  select(-moe)
```

## Practice some tidyverse functions for data wrangling

- `filter()`   Use for rows!
- `select()`   Use for columns!
- `mutate()`    already existing column
- `case_when()`
- `group_by() %>% summarize()`

### dplyr::filter() to keep or exclude rows based on our conditions

1. Exact match (single condition)

only keep observations where the island is Biscoe
```{r}
ex_1 <- filter(.data = penguins, island == "Biscoe")
unique(ex_1$island)
```
2. "and" statement within filter

only keep observations if the penguin species is Chinstrap AND the year is 2008
```{r}
ex_2 <- penguins %>%  filter(species == "Chinstrap", year == 2008)
```

3. Exclude Gentoos

keep observations if the penguin species is not Gentoo
```{r}
ex_3 <- penguins %>% filter(species != "Gentoo")
```

Create a subset from penguins that only contains observations for Adelie penguins with a flipper length greater than or equal to 195mm

```{r}
practice_1 <- penguins %>% filter(species == "Adelie", flipper_length_mm >= 185)
```

4. Using the %in% operator

Use %in% operator to look for any matches within a vector, and return TRUE if the value does exist for our row
we want to keep observations where the island is Dream or Torgersen
```{r}
ex_4 <- penguins %>% filter(island %in% c("Dream", "Torgersen"))
```

### select() to select or exclude COLUMNS

5. Keep a single column by name

only want to keep the species column
```{r}
ex_5 <- penguins %>%  select(species)
ex_5 <- select(.data = penguins, species)
```

6. Keep multiple columns by name

create a subset that contains species, bill_length_mm through flipper_length_mm, and year
```{r}
ex_6 <- penguins %>% select(species, bill_length_mm : flipper_length_mm, year)
```

7. Choose or exlclude columns based on position
```{r}
ex_7 <- penguins %>% 
  select(-4)
```

8. A piped sequence 

write a piped sequence where you first only keep columns for species, body_mass_g, and island, and second: filter to only include observations if body_mass_g is greater than 3600

```{r}
ex_8 <- penguins %>% 
  select(species, body_mass_g, island) %>% 
  filter(body_mass_g > 3600)
```

### mutate() to add or update existing colums, while keeping the ones already there

9. add a column that has body mass converted to kg

```{r}
ex_9 <- penguins %>% 
  mutate(body_mass_kg = body_mass_g / 1000) %>% 
  mutate(flipper_length_m = flipper_length_mm / 1000)
```

10. Mutate to update and existing column

```{r}
ex_10 <- penguins %>% 
  mutate(island = str_replace(island, pattern = "sen", replacement = "HOORAY!"))
```

11. Update the class of the bill depth column to character

```{r}
ex_11 <- penguins %>%
  mutate(bill_depth_mm = as.character(bill_length_mm))
```

### case_when(): your favorite if-else
Create a new column that contains "small" if penguin flipper length is less than 185, medium otherwise
```{r}
ex_12 <- penguins %>% 
  mutate(binned_size = case_when(
    flipper_length_mm < 185 ~ "small", 
    flipper_length_mm >= 185 ~ "medium"
  )
  )
```

Task: starting from penguins, create a new column called bird_bill_bin that contains:
- if bill length < 38 mm: "short bill"
- if bill length >= 38mm and < 43 mm: "medium bill" 
- if bill length >= 43: "long bill"

```{r}
practice_12 <- penguins %>% 
  mutate(bird_bill_bin = case_when(
    bill_length_mm < 38 ~ "short bill",
    bill_length_mm >= 38 & bill_length_mm < 43 ~ "medium bill",
    TRUE ~ "long bill" #includes NA values
    
  ))
```

### group_by() %>% summarize() to find summary values by group, and return them in a nice summary table

```{r}
mean(penguins$bill_depth_mm, na.rm = TRUE)

```
13. group_by() to say there are groups within the species column

```{r}
ex_13 <- penguins %>% 
  group_by(species, island) %>% 
  summarize(mean_flipper_length = mean(flipper_length_mm, na.rm = TRUE),
            sd_flipper_length = sd(flipper_length_mm, na.rm = TRUE))
```

## practice

only keep penguins that have a body mass > 3500 g
only keep colums for species, flipper_length_mm and sex
create a summary table that contains the mean, sd, max and min value of flipper length by species and sex (as the grouping variables)

```{r}
practice_a <- penguins %>% 
  filter(body_mass_g > 3500) %>% 
  select(species, flipper_length_mm, sex) %>% 
  group_by(species, sex) %>% 
  summarize(mean_flipper_length = mean(flipper_length_mm, na.rm = TRUE),
            sd_flipper_length = sd(flipper_length_mm, na.rm = TRUE),
            max_flip = max(flipper_length_mm, na.rm = TRUE),
            min_flip = min(flipper_length_mm, na.rm = TRUE)) %>% 
  ungroup() %>% 
  drop_na(sex)
```

